---
title: "A Forecast of U.S. Solar Energy Generation"
author: "David Noonan"
date: "`r Sys.Date()`"
output:
  html_document:
    code_download: true
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---
<center>[Made with R Markdown.  See the code on my GitHub Page](https://github.com/DavidNoonan/Forecasting-Energy-Generation)</center> 

<br>

This is a timeseries analysis of the total solar electricity generation in the U.S. based on data from the U.S. Energy Information Administration. 
The data contains monthly totals from 1973-2017 by energy source, with units in quadrillions of British Thermal Units. I converted the units to gigawatt-hours, which I think is more common for electricity.

My goal is to fit an ARIMA timeseries model, and make 12-month and 5-year forecasts. This isn't a rigorous study based on extensive domain knowledge. That would certainly help, but for now I am focusing on modeling. This is an exercise in building an ARIMA model. Before we dive into model building, here is a plot of all the other sources including solar, for some context.

```{r, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(dpi = 300)

# load packages
library(tidyverse) # Tidyverse + lubridate for data manipulation
library(lubridate) # package for handling time data
library(tm) # text mining package
library(astsa) # time series modeling package
library(gghighlight) # to highlight in plots
```



```{r, message=FALSE, warning=FALSE, include=FALSE}
# obtain data from eia.gov website. This dataset contains electricity generation
# monthly for the total U.S., categorized by source.

total_energy_data <- read.csv(
  file = "https://www.eia.gov/totalenergy/data/browser/csv.php?tbl=T01.02",
  stringsAsFactors = FALSE
)

qbtu_to_gigawatthours <- 293071.070172222215 # ratio of gigawatthours per quadrillion btus

# data cleaning
cleaned_energy_data <- total_energy_data %>%
  mutate(Description = str_remove_all(Description, " Production| Energy| Power| Electric| \\(Dry\\)| Plant Liquids"),
         Value = ifelse(Value == "Not Available", NA, Value) %>% as.numeric(),
         gwh = as.numeric(Value) * qbtu_to_gigawatthours, # convert btus to gigawatthours
         twh = gwh/1000, # terawatt hours
  YYYYMMDD = ymd(paste(YYYYMM, "01", sep = "")) # convert YYYYMM to date + first day of the month
         ) %>%
  filter(is.na(YYYYMMDD) == FALSE) %>%
  group_by(Description, YYYYMMDD) %>%
  summarize(gwh = sum(gwh),
            twh = sum(twh)) %>%
  ungroup()


```


```{r, message=FALSE, warning=FALSE, include=FALSE}
head(cleaned_energy_data) # look at structure of total_energy_data dataframe

# examine energy sources
(cleaned_energy_data_sources <- unique(cleaned_energy_data$Description))

# look at the range of dates
range(cleaned_energy_data$YYYYMMDD, na.rm = TRUE)

# look at the range of values for generation
range(cleaned_energy_data$Value, na.rm = TRUE)
```

```{r echo=FALSE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE, paged.print=FALSE}
# plot generation sources over time
(sourcesplot <- cleaned_energy_data %>%
  filter(
    stringr::str_detect(string = Description, pattern = "Total") == FALSE,
    Description %in% c("Natural Gas", "Crude Oil", "Coal", "Nuclear", "Solar", "Wind")
  ) %>%
  ggplot() +
  geom_path(
    mapping = aes(
      x = YYYYMMDD,
      y = twh,
      color = Description
    ),
    na.rm = TRUE,
    size = .72
  ) +
  ggtitle(label = paste0("U.S. Electricity Generation 1973-", year(max(cleaned_energy_data$YYYYMMDD)), " by Source")) +
  xlab("Year") +
  ylab("Generation (teraawatt-hours)") +
  guides(colour = guide_legend(override.aes = list(size = 5))) +
  gghighlight(Description == "Solar") +
  geom_text(
    data = cleaned_energy_data %>% filter(
      YYYYMMDD == dplyr::last(YYYYMMDD),
      stringr::str_detect(string = Description, pattern = "Total") == FALSE,
      Description %in% c("Natural Gas", "Crude Oil", "Coal", "Nuclear", "Solar", "Wind")
    ),
    aes(
      label = Description,
      x = YYYYMMDD + 0.5,
      y = twh
    )
  ) +
  theme_minimal() +
  theme(legend.position = "none"))
```



If we look carefully, solar energy is one of the smallest sources, shown as a line on the bottom right of the plot. It is dwarfed by the major energy sources: natural gas, coal, and oil. Nevertheless, solar is experiencing exponential growth. We take a closer look at solar and wind, two renewable sources, in the next plot.

```{r echo=FALSE, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}


(solar_plot <- cleaned_energy_data %>% filter(Description %in% c("Solar"),
                                                      is.na(gwh) == FALSE,
                                                      year(YYYYMMDD) >= 1990) %>%
  ggplot() +
  geom_path(
    mapping = aes(
      x = YYYYMMDD,
      y = gwh,
      color = Description
    ),
    na.rm = TRUE,
    size = .72
  ) +
  ggtitle(label = paste0("U.S. Solar and Wind Generation 1990-", year(max((cleaned_energy_data$YYYYMMDD))))) +
  xlab("Year") +
  ylab("Gigawatt-hours (thousands)") +
  scale_y_continuous(labels = function(x) x/1000) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  labs(color = "Source")) +
  ggthemes::theme_tufte() +
  theme(legend.position = "none")
```


Solar energy generation is growing exponentially. How much will it grow in the near future? Here we will make a forecast with an ARIMA model.
  
  
First we need to do some transformations on the data. ARIMA requires a "weakly stationary" series, which assumes no trend, and constant variance. In the case above, the variance appears to increase with time. We can force a constant variance by applying a logarithmic transformation to generation. This means that we are modeling the natural logarithm of generation, rather than generation itself. Below is a plot of the log transformed solar generation:  
  
```{r, echo=FALSE, message=FALSE, warning=FALSE}


(solar_and_wind_plot <- cleaned_energy_data %>% filter(Description %in% c("Solar"),
                                                      is.na(gwh) == FALSE,
                                                      year(YYYYMMDD) >= 2010) %>%
  ggplot() +
  geom_path(
    mapping = aes(
      x = YYYYMMDD,
      y = log(gwh),
      color = Description
    ),
    na.rm = TRUE,
    size = .72
  ) +
  ggtitle(label = paste0("U.S. Solar Energy Generation 1990-", year(max((cleaned_energy_data$YYYYMMDD))))) +
  xlab("Year") +
  ylab("Gigawatt-hours (thousands)") +
  labs(color = "Source")) +
  ggthemes::theme_tufte() +
  theme(legend.position = "none")

```

The result is a nice constant trend that oscillates with the seasons. We can't model this with ARIMA yet, because it has a trend. ARIMA models assume a "stationary process" (the mean and variance do not change over time). Fear not, we can remove the trend with another transformation--differencing, where the previous value is subtracted from each observation. This is reversible so long as we know the first observation. Below is the differenced series:

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# make solar energy table with logged and differenced-logged generation columns
cleaned_solar_energy_data <- cleaned_energy_data %>% filter(Description %in% c("Solar"),
                                                      year(YYYYMMDD) >= 2010) %>%
  mutate(log_gwh = log(gwh),
         diff_log_gwh = log_gwh - lag(log_gwh))



(difflogsolarplot <- cleaned_solar_energy_data %>% ggplot() +
  geom_path(
    mapping = aes(
      x = YYYYMMDD,
      y = diff_log_gwh,
      color = Description
    ),
    na.rm = TRUE,
    show.legend = FALSE,
    size = .7
  ) +
  ggtitle(label = "U.S. Solar Generation (differenced log transform) 2010-2017") +
  xlab("Year") +
  ylab("Generation log(gigawatt-hours)") +
  ggthemes::theme_tufte()+
  theme(legend.position = "none"))
```

Now it looks like a stationary process, great for fitting ARIMA. To begin, We need to come up with candidate parameters. We can find those by looking at auto-correlation (ACF) and partial-auto-correlation (PACF) plots. The parameters in question are the auto-regressive (AR), moving average (MA), and seasonal parameters. For details on this, I refer to *Time Series Analysis and its Applications* by Robert H. Shumway and David S. Stoffer.  
  
  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# create timeseries object of solar generation to interface with ASTSA functions
solar_ts <- ts(cleaned_solar_energy_data$gwh, start = c(2010, 1), end = c(2017, 6), frequency = 12)

# log and diff transformation of the timeseries object above
log_diff_solar_ts <- diff(log(solar_ts))


# change graphical parameters to show less white space
invisible((acfpar <- par(mfrow = c(2, 1), mar = c(2, 4, 1, 2), xlab = NA)))
# autocorrelation function plot
acf(log_diff_solar_ts, xlab = NA, main = NA, lag.max = 50)
# partial autocorrelation plot
pacf(log_diff_solar_ts, main = NA, lag.max = 50)
```

  
  In the above ACF plot, we see a pattern of strong peaks at lag years 1, 2, 3... slowly decaying. This may indicate some seasonal non-stationarity, which could be fixed with seasonal differencing. We will consider this in subsequent models.  
  In the PACF plot, we see strong autocorrelation at the 12 month lag (1.0 on the plot), which cuts off, indicating a seasonal ARIMA model of order P = 1 may be helpful.  Below, we fit an ARIMA model with order P = 1, with differencing D = 0 and d = 1:  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# create log transformation of solar time series object
logsolar <- log(solar_ts)
# fit ARIMA model order d = 1, D = 0, P = 1
fit1 <- sarima(logsolar, 0, 1, 0, P = 1, D = 0, Q = 0, S = 12, details = FALSE)
```
```{r, echo=FALSE, fig.height=2, fig.width=7, message=FALSE, warning=FALSE}
# graphical parameters to reduce white space
invisible((acfpar <- par(mfrow = c(1, 1), mar = c(2, 4, 1, 2))))

# pacf plot of residuals of model "fit1"
pacf(fit1$fit$residuals, lag.max = 50, xlab = NA)
```


The diagnostic plots are encouraging.  The residual time plot shows no obvious patterns, with no outliers greater than 3 standard deviations from zero. The residual distribution in the Q-Q plot is very close to normal, despite heavy tails. Two concerns are the residual ACF plot showing autocorrelation in the 12th lag, which indicates a seasonal moving average component of order Q = 1 may be helpful. The PACF plot shows a peak at lag 12, so we may also consider another seasonal auto-regressive order like P = 2. Finally, the p-values of the Ljung-Box statistics indicate some departure from the independence assumption of the residuals.  

Next, we will compare a series of models using information criterea:  
Fit 3: ARIMA$(0,1,0) \times (1,0,1)_{12}$  
Fit 4: ARIMA$(0,1,0) \times (1,1,1)_{12}$  
Fit 5: ARIMA$(0,1,0) \times (2,1,1)_{12}$  
Fit 6: ARIMA$(1,1,0) \times (1,1,1)_{12}$  
Fit 7: ARIMA$(1,1,1) \times (1,1,1)_{12}$  
Fit 8: ARIMA$(0,1,1) \times (1,1,1)_{12}$  

```{r, message=FALSE, warning=FALSE, include=FALSE}
# fit arima model candidates :
fit3 <- sarima(logsolar, 0, 1, 0, P = 1, D = 0, Q = 1, S = 12, details = FALSE)
fit4 <- sarima(logsolar, 0, 1, 0, P = 1, D = 1, Q = 1, S = 12, details = FALSE)
fit5 <- sarima(logsolar, 0, 1, 0, P = 1, D = 1, Q = 1, S = 12, details = FALSE)
fit6 <- sarima(logsolar, 1, 1, 0, P = 1, D = 1, Q = 1, S = 12, details = FALSE)
fit7 <- sarima(logsolar, 1, 1, 1, P = 1, D = 1, Q = 1, S = 12, details = FALSE)
fit8 <- sarima(logsolar, 0, 1, 1, P = 1, D = 1, Q = 1, S = 12, details = FALSE)

# fit4 <- sarima(logsolar, 0,1,0,P = 2, D = 0, Q = 1, S = 12,details = FALSE)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# create vector of AICs
fitaics <- c(fit3$AIC, fit4$AIC, fit5$AIC, fit6$AIC, fit7$AIC, fit8$AIC)
fitnames <- c("fit 3", "fit 4", "fit 5", "fit 6", "fit 7", "fit 8")
dotchart(fitaics, labels = fitnames, main = "Model AIC comparison", ylab = "ARIMA Model", xlab = "AIC")
```

We see in the dot chart above that fit 8 has the smallest AIC value. Fit 8 is ARIMA$(0,1,1) \times (1,1,1)_{12}$.  Below is a plot of its diagnostics:  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# fit5 <- sarima(logsolar, 0,1,1,P = 1, D = 1, Q = 1, S = 12,details = FALSE)
```

The diagnostic plots look reasonable for this model, with the exception of two outliers among the standardized residuals. With that kept in mind, We will move forward with this model because it looks good enough. As a diagnostic, below is a 12 month forecast of the mean, using the last 12 months as hold out.

```{r, message=FALSE, warning=FALSE, include=FALSE}
# make holdout indices
holdout <- (length(logsolar) - 11):length(logsolar)

# create timeseries object with hold out year
logsolarholdout <- logsolar[-holdout]

# forecast
holdforc <- sarima.for(logsolarholdout,
  n.ahead = 13, p = 0, d = 1, q = 1, P = 1,
  D = 1, Q = 1, S = 12
)

# build dataframe of forecast with holdout
holdforcdf <- data.frame(
  prediction = c(
    rep(NA, length(cleaned_solar_energy_data$YYYYMMDD) - 13),
    exp(holdforc$pred)
  ),
  date = cleaned_solar_energy_data$YYYYMMDD
)
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot generation sources over time
logsolarplot <- ggplot(data = cleaned_solar_energy_data, aes(x = holdforcdf$date)) +
  geom_path(
    mapping = aes(
      x = cleaned_solar_energy_data$YYYYMMDD,
      y = cleaned_solar_energy_data$gwh,
      color = "Actual"
    ),
    na.rm = TRUE,
    show.legend = FALSE,
    size = .7
  ) +
  ggtitle(label = "Forecast of Solar Energy Generation (holdout period)") +
  xlab("Year") +
  ylab("Generation (gigawatt-hours)") +

  # holdoutpred <- ggplot(data = holdforcdf) +

  geom_path(
    mapping = aes(
      x = holdforcdf$date,
      y = holdforcdf$prediction,
      color = "Predicted"
    ),
    na.rm = TRUE,
    size = 1
  ) +
  geom_vline(xintercept = as.numeric(as.Date("2016-06-01")), linetype = 2)

logsolarplot + labs(color = "") + ggthemes::theme_tufte()
```
  
The model predicts the holdout period pretty well!  
  
Now we can can extrapolate to the actual future, and forecast the energy output 5 years out from June 2017:

```{r, message=FALSE, warning=FALSE, include=FALSE}
# 5-year forecast
forc5 <- sarima.for(logsolar,
  n.ahead = 60, p = 0, d = 1, q = 1, P = 1,
  D = 1, Q = 1, S = 12
)

# create vector of corresponding dates for the forecast
forc5dates <- c(cleaned_solar_energy_data$YYYYMMDD, seq(as.Date("2017-07-01"),
  as.Date("2022-06-01"),
  by = "months"
))

# create dataframe for the 5 year forecast with 95% confidence intervals
forc5df <- data.frame(
  prediction = c(
    rep(NA, length(cleaned_solar_energy_data$gwh)),
    exp(forc5$pred)
  ),
  UL = c(
    rep(NA, length(cleaned_solar_energy_data$gwh)),
    exp(forc5$pred + qnorm(0.975) * forc5$se)
  ),
  LL = c(
    rep(NA, length(cleaned_solar_energy_data$gwh)),
    exp(forc5$pred - qnorm(0.975) * forc5$se)
  ),
  date = forc5dates,
  gwh = c(
    cleaned_solar_energy_data$gwh,
    rep(NA, 60)
  )
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# plot generation sources over time
forecast5plot <- ggplot(data = forc5df) +
  geom_path(
    mapping = aes(
      x = forc5df$date,
      y = forc5df$gwh,
      color = "Actual"
    ),
    na.rm = TRUE,
    show.legend = FALSE,
    size = .7
  ) +
  ggtitle(label = "5 Year Forecast of Solar Energy Generation") +
  xlab("Year") +
  ylab("Generation (gigawatt-hours)") +

  # holdoutpred <- ggplot(data = holdforcdf) +

  geom_path(
    mapping = aes(
      x = forc5df$date,
      y = forc5df$prediction,
      color = "Predicted"
    ),
    na.rm = TRUE,
    size = 1
  ) +
  geom_vline(xintercept = as.numeric(as.Date("2017-06-01")), linetype = 2) +
  geom_ribbon(aes(x = date, ymin = LL, ymax = UL, fill = "95% conf."), alpha = 0.75) +
  scale_fill_manual("", values = "grey")

forecast5plot + labs(color = "") + ggthemes::theme_tufte()
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
# 95% confidence interval for solar generation year 2022
(conf2022 <- c(forc5df$LL[length(forc5df$date)], forc5df$UL[length(forc5df$date)]))

# 95% confidence interval for solar generation year 2018
(conf2018 <- c(forc5df$LL[which(forc5df$date == as.Date("2018-06-01"))], forc5df$UL[which(forc5df$date == as.Date("2018-06-01"))]))

# generation in June 2017
june2017solar <- cleaned_solar_energy_data$gwh[length(cleaned_solar_energy_data$YYYYMMDD)]

# percent difference in 2022
(percent_diff_2022 <- (conf2022 / june2017solar - 1) * 100)

# percent difference in 2018
(percent_diff_2018 <- (conf2018 / june2017solar - 1) * 100)
```

As we see above, our ARIMA model seems to provide a believable forecast for the growth of solar energy generation. It predicts a continuation of exponential growth. This should be taken cautiously, because solar energy will likely experience changes in growth as our energy infrastructure changes from fossil fuel sources. In the near term however, I hope the model is accurate. 
 
By June of the year 2022, the model predicts with 95% confidence that the energy generation will be between 78341.97 and 203573.46 gigawatt-hours, which is a 200 to 687 percent increase. In 2018, one year out from the end of the data, the model predicts between 29448.10 to 42313.16 gigawatt-hours, which is a 13.8 to 63.6 percent increase from a year before.
 
Thanks for reading, and thanks to Prof. Kerr of CSU East Bay for introducing me to this stuff.
